---
title: "Ecology-aware MkPrime: mathematical details"
format: html
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Ecology-aware MkPrime: mathematical details}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library("MkPrime")
```

## Motivation

The Mk model [@Lewis2001] and its extensions treat every morphological
character as an independent draw from a shared substitution process. In
reality, distantly related lineages that come to occupy the same
ecological niche tend to evolve similar morphologies: rodents that
adopt fossorial habits, fish that adopt deep-water habits, plants that
adopt epiphytic habits, all converge on a recognisable phenotype.
Under the independence assumption, the resulting parallel
character-state changes look like phylogenetic signal, and the
inferred tree groups together lineages by ecology rather than by
descent.

Ecology-aware MkPrime (enabled via `MkPrimeModel(ecologyAware = TRUE)`)
addresses this by letting each character's substitution rate on each
edge depend on the ecology assigned to that edge. The user supplies a
discrete ecology coding at the tips of the tree (e.g. "terrestrial /
arboreal / fossorial"). The model then infers, for each character and
each ecology, whether transitions are *encouraged* (faster than
baseline), *discouraged* (slower than baseline), or unaffected, and by
how much. Convergent characters can be downweighted on the edges of
the tree where convergence is plausible, returning the phylogenetic
signal to its rightful place in the rest of the matrix.

## The model

### Edge ecology partition

Tip ecologies are supplied as integer codes
$y_t \in \{0, \ldots, K - 1\}$ for each tip $t$, with $K \geq 2$. We
treat the ecology as a single $K$-state character evolving on the tree
under JC($K$), reusing the existing pruning machinery for transformational
characters. Branch lengths and (optionally) ACRV are shared with the
morphological likelihood.

A forward pruning pass on the ecology character computes the
conditional likelihood at every internal node. A single backward pass
then yields the per-node marginal posterior
$P(\text{node}_n = e \mid \text{tree}, \text{ecology data})$ at every
internal node, for every ecology state $e \in \{0, \ldots, K - 1\}$.
For each edge $p \to c$, we define the **edge-ecology weights**

$$
w_{p \to c}(e) \;:=\; P(\text{node}_p = e \mid \text{tree}, \text{ecology data}),
$$

with the root edge using the stationary distribution $1/K$. The
weights form an $n_{\text{edge}} \times K$ matrix that is recomputed
once per likelihood evaluation and cached.

### Reference ecology and identifiability

With $K$ ecology states, only $K - 1$ degrees of freedom in the
rate-modifier vector are identifiable from the data: the rate factor
"in ecology $A$" and "in ecology $B$" are complementary, not
independent. We therefore dummy-code against a **reference ecology**
$e_{\text{ref}}$, chosen automatically as the ecology with the
largest edge-weighted mass in the data; the reference ecology gets
rate factor 1 by construction, and only non-reference ecologies carry
a latent $z$.

The choice of reference is mathematically arbitrary -- a different
choice gives an equivalent posterior up to a sign flip on $\log \varphi$
and the appropriate relabelling of $\theta_e$. We pick the most-frequent
ecology because that loads most of the rate information into the
well-conditioned $z = 0$ "no deviation" anchor.

### Per-character rate modification

For each character $c$ and each NON-reference ecology
$e \in \{0, \ldots, K-1\} \setminus \{e_{\text{ref}}\}$ we introduce
a latent

$$
z_{c, e} \in \{\text{none}, \text{encouraged}, \text{discouraged}\},
$$

with the asymmetric spike-and-slab prior

$$
P(z_{c, e} = \text{none}) = \pi_0, \quad
P(\text{enc}) = (1 - \pi_0)\,\theta_e, \quad
P(\text{disc}) = (1 - \pi_0)(1 - \theta_e).
$$

The hyperpriors are
$\pi_0 \sim \mathrm{Beta}(7, 3)$ (mean $0.7$),
$\theta_e \sim \mathrm{Beta}(\alpha_\theta, \beta_\theta)$ (default
$\mathrm{Beta}(2, 2)$, weakly symmetric but allowing biological
asymmetry between enhancement and depression), and
$\log \varphi \sim \mathcal{N}(0, 0.5^2)$.

### Rate normalisation by $\gamma$

The crucial step. Without normalisation, the prior-expected per-edge
rate factor on a non-reference ecology is

$$
\gamma_e(\pi_0, \theta_e, \varphi) \;=\; \pi_0 + (1 - \pi_0)\left[\theta_e\,\varphi + (1 - \theta_e)\,\varphi^{-1}\right] \;\geq\; 1,
$$

with equality iff $\varphi = 1$. Switching the ecology layer on then
inflates the prior-expected rate by $\gamma_e$, and the chain
compensates by stretching the tree length by $1/\gamma_e$. Three
rate-related parameters
(`tree_length`, baseline character rates, $\varphi$) lie on a single
identifiability ridge, and the MCMC follows it badly.

The ACRV-style fix: divide every rate factor by $\gamma_e$ so the
prior-expected factor per non-reference cell is exactly 1. The
applied modification then becomes, for a non-reference ecology $e$,

$$
\tilde\mu(z_{c, e}, \varphi, \pi_0, \theta_e) =
\frac{\mu(z_{c, e}, \varphi)}{\gamma_e(\pi_0, \theta_e, \varphi)},
\qquad
\mu(z, \varphi) = \begin{cases}
  1 & z = \text{none} \\
  \varphi & z = \text{enc} \\
  1/\varphi & z = \text{disc}
\end{cases}
$$

with the convention that the reference ecology has rate factor 1
unconditionally. The transition matrix for character $c$ on an edge
$p \to c$ with branch length $t$ is the mixture

$$
P_c(t) = \sum_{e = 0}^{K-1} w_{p \to c}(e) \cdot
  P_c^{(e)}\bigl(t \cdot \tilde\mu_{c, e}\bigr),
$$

with $\tilde\mu_{c, e_{\text{ref}}} \equiv 1$.

For a **transformational** character $c$ this is the same scaling
applied symmetrically to all transitions; for a **neomorphic**
character (gain rate $r_{01}$, loss rate $r_{10}$) the
encouraged-direction simultaneously multiplies gains by
$\varphi/\gamma_e$ and divides losses by $\varphi/\gamma_e$,
preserving the gain/loss asymmetry that produces state-level
convergence between unrelated lineages sharing an ecology.

### Why z is identifiable

Two observations:

1. **Transformational characters are symmetric in state labels**, but
   $z = \text{enc}$ (rate factor $\varphi$) and $z = \text{disc}$
   (rate factor $1/\varphi$) produce different *amounts* of change
   over a given branch length and are therefore distinguishable from
   the data, even though the off-diagonal pattern of the rate matrix
   is unchanged.

2. **Neomorphic characters are never relabelled**: 0 always means
   absent and 1 always means present. The gain rate and loss rate
   are physically distinct quantities and not interchangeable. So
   $z = \text{enc}$ (more gains, fewer losses) and $z = \text{disc}$
   (fewer gains, more losses) leave a directional signature that is
   identifiable from the proportion of state-1 tips per clade and the
   shape of the per-clade subtree pattern.

The reference-ecology dummy-coding combined with full
$\gamma$-normalisation closes the rate--time ridge, removes the
$K$-vs-$K{-}1$ redundancy, and lifts the symmetric-slab assumption
that the prior commits to equal numbers of enhanced and depressed
characters.

### Likelihood

The full character-likelihood factorises as

$$
\log L \;=\; \sum_{c} \log L_c,
$$

with each $L_c$ computed by Felsenstein pruning using the
ecology-mixture transition matrix above. The mixture is taken inside
the edge loop, so the per-character cost scales as
$O(n_{\text{edge}} \cdot K \cdot k^2)$ instead of the usual
$O(n_{\text{edge}} \cdot k^2)$: a factor of $K$ overhead.

## Caveats and edge cases

* **Reference-ecology choice is not data-driven beyond edge mass.**
  Picking the most-frequent ecology as reference is a numerical
  convenience, not a model assumption -- a different choice gives
  the same posterior on the tree and on quantities of interest
  ($|\log \varphi|$, $\pi_0$, the per-character pattern of
  affected vs unaffected ecologies), modulo relabelling. The
  default is good unless the most-frequent ecology is also
  biologically anomalous (e.g. an artificially aggregated
  "unknown" category), in which case override
  `referenceEcology` in `MkPrimeData`.

* **Skewed edge-ecology degeneracy.** Even with $\gamma$-normalisation,
  if essentially every edge in the tree carries the same ecology
  (e.g. $26/30$ in some simulated test cases) there is little
  information left to identify $\varphi$: the model can absorb the
  effect into baseline rates. $\pi_0$ remains identifiable, but
  $|\log \varphi|$ does not. Real data typically distribute ecology
  more evenly along the backbone.

* **Sparse-cell estimation.** Individual $z_{c, e}$ posteriors are
  weakly identified at typical morphological matrix sizes
  (~200 characters $\times$ a few tip-states-worth of edges).
  Rank the top effects with confidence; treat the long tail with
  appropriate uncertainty. The aggregate $\pi_0$ posterior is much
  more robust than any individual cell.

## MCMC

The ecology layer adds the following moves to the existing MkPrime
proposal scheme:

* **Bactrian scale** on $\log \varphi$ (or each $\log \varphi_e$),
  reusing the kernel at `src/mcmc.cpp`. Under v2 this move now has a
  likelihood contribution via $\gamma_e$ on every non-reference edge
  (it did not in v1).
* **Logit random walk with Jacobian** on $\pi_0$ and on each
  $\theta_e$ (one per non-reference ecology). Both now carry a
  likelihood contribution via $\gamma_e$.
* **Gibbs sweep** over each $(c, e)$ cell of the $z$ matrix every
  $K_{\text{sweep}}$ generations, for $e \neq e_{\text{ref}}$: enumerate
  the three values $\{0, 1, 2\}$, compute the conditional
  log-posterior of each under the asymmetric slab, and sample
  categorically.

All four moves are integrated into the cost-adjusted move-weight
scorer; the Gibbs sweep is the cheapest per-step (no likelihood
recompute beyond the per-cell change) and the others recompute the
full ecology-mixed likelihood. The Gibbs sweep's weight is pinned so
the scorer does
not preferentially inflate it.

## Simulation evidence

Three validation simulations live under `inst/simulations/ecology/`.
Each is a self-contained R script that simulates from
`.SimulateMkPrimeEcology()` and runs MCMC under the ecology-aware
model.

### Sim 1 -- null

A random 16-tip tree with random binary tip ecology, 120 characters
simulated with $z = 0$ everywhere. Both hyperparameters land where
the model says they should under $H_0$: $\pi_0$ posterior mean
$\approx 0.68$ (CI $[0.51, 0.84]$, prior mean $0.7$); $\varphi$
posterior mean $\approx 1.04$ (CI $[0.16, 1.69]$, centred at $1$).
The ecology layer does not introduce spurious signal when ecology is
irrelevant.

### Sim 2 -- recovery

A random 16-tip tree with random binary tip ecology, with half of the
transformational characters simulated under $z = \text{encouraged}$ in
ecology 1 (and $z = 0$ in ecology 0) and $\varphi = 3$. The result
on $\pi_0$ is clean ($\pi_0$ posterior $\approx 0.5$, well below the
prior mean $0.7$, correctly detecting that $\approx$ 50% of cells
carry a non-zero effect). The result on $\varphi$ in this regime is
the rate-vs-ecology degeneracy described above: with edge ecology
heavily skewed toward ecology 1 (the random tip ecology happened to
yield $26/30$ edges in ecology 1), the model absorbs the effect into
baseline rates and the $\varphi$ posterior remains near $1$. This is
a sim-design issue, not a model bug; it documents the regime in which
$\varphi$ is hard to identify even when $\pi_0$ is.

### Sim 3 -- convergent ecology

The headline demonstration. Sixteen tips arranged as four 4-tip
clades $A, B, C, D$, with the true topology
$((A, C), (B, D))$ but ecology shared by $A$ and $B$ (and separately
by $C$ and $D$). Sixty neomorphic characters are simulated under
$z = \text{encouraged}$ in ecology 1, generating genuine convergent
state changes on the $A$ and $B$ edges; 180 transformational
characters carry the baseline phylogenetic signal. Branch lengths are
tuned (`stemBranch = 0.10`, `rootBranch = 0.15`) so that on baseline
data alone the true tree wins under parsimony, while on the full
convergent dataset the convergence-driven tree $((A, B), (C, D))$
beats the truth under parsimony in 6--8 of 8 pilot replicates.

On the 200{,}000-generation MCMC of a single replicate at the
Goldilocks config:

* **Hyperparameters detect the effect.** The aware-model posterior
  on $\pi_0$ concentrates at $\approx 0.38$ (CI $[0.22, 0.72]$),
  well below the prior mean $0.7$, and on $\varphi$ at $\approx 7$
  (CI $[1.0, 32]$), with the CI excluding $1$. The model correctly
  detects that ecology is exerting a non-trivial effect on the
  characters.
* **Tree-topology posterior is broad in both chains.** Across 200
  post-burn-in samples, neither the blind nor the aware chain places
  appreciable posterior probability on either reference bipartition.
  Mean CID-to-truth and CID-to-wrong are essentially tied for both
  chains, with a slight bias toward truth (CID gap $\approx 0.01$
  blind, $\approx 0.005$ aware). The chain has mixed sufficiently to
  escape the parsimony-tree starting basin, and the posterior over
  topologies for this dataset is genuinely broad rather than
  concentrated on either reference.

In the **negative control** Sim 3b ($z = 0$ everywhere, so $A$ and
$B$ share ecology by coincidence not by mechanism), the aware-model
posterior correctly stays near the prior: $\pi_0 \approx 0.79$ (prior
mean $0.7$), $\varphi \approx 1.4$ (CI includes $1$). The ecology
layer does not invent an effect where there is none.

The headline conclusion at the level of identifiable summaries
($\pi_0, \varphi$) is therefore clean: the ecology-aware model
turns on for genuinely convergent data and stays off for the matched
null. At the level of tree-topology recovery, the 200k-iter chain on
240 characters explores too broad a posterior to lock onto either
reference tree, and further work is needed -- multi-replicate runs
with longer chains, tighter tree-length priors, or larger character
counts -- to obtain a clean topology-level demonstration.

## Usage

```{r usage}
#| eval: false
# Tip ecology as a named integer vector, aligned to mkd$taxon_names.
ecology <- c(T1 = 0L, T2 = 0L, T3 = 1L, T4 = 1L, ...)

mkd  <- MkPrimeData(myMatrix,
                    neomorphic = neoIdx,
                    ecology    = ecology)
# mkd$refEcology now holds the auto-selected reference (most-frequent
# ecology by tip count, refined to edge mass at MCMC start).

model <- MkPrimeModel(
  ecologyAware  = TRUE,
  magnitudeMode = "global",   # or "per_ecology"
  rho0Alpha     = 7,  rho0Beta = 3,   # Beta prior on pi0
  thetaAlpha    = 2,  thetaBeta = 2,  # Beta prior on each theta_e
  sigmaPhi      = 0.5
)

res <- RunMkPrime(mkd, model = model,
                  mcmc = MkPrimeMCMC(nIter = 1e5))
```

After the run, the per-character / per-ecology posterior on $z$ is
available in the chain output, alongside the standard topology,
branch-length and rate samples. Visualisations of the
$z$ matrix highlight which character/ecology cells contribute to the
convergence signal.

## References
