---
title: "Q-matrix heterogeneity: mathematical details"
format: html
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Q-matrix heterogeneity: mathematical details}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(MkPrime)
```

## Motivation

The Mk model [@Lewis2001] for discrete morphological characters assumes
equal equilibrium state frequencies: every character with $k$ observed
states uses the Jukes--Cantor rate matrix with $\pi_j = 1/k$ for all $j$.
This is a strong assumption. In practice, some characters may have
one dominant state while others are more evenly distributed. If the
data contain this kind of variation and the model ignores it, branch
lengths and topologies can be biased.

Q-matrix heterogeneity (enabled via `MkPrimeModel(qHeterogeneity = TRUE)`)
relaxes the equal-frequency assumption by integrating each character's
likelihood over a mixture of F81 rate matrices [@Felsenstein1981] with
varying equilibrium frequencies. The mixture is controlled by a single
MCMC-sampled parameter, `beta_scale` ($= \alpha$), which acts as the
concentration of a symmetric Dirichlet distribution.

## The model

### F81 transition probabilities

For a character with $k$ states and equilibrium frequencies
$\boldsymbol{\pi} = (\pi_0, \ldots, \pi_{k-1})$, the F81 model gives
transition probabilities in closed form:

$$
P_{ij}(t) = \pi_j + (\delta_{ij} - \pi_j) \exp(-\mu t)
$$

where $\delta_{ij}$ is the Kronecker delta and $\mu$ is the total
substitution rate (normalised so that the expected number of changes per
unit time is 1):

$$
\mu = \frac{1}{1 - \sum_j \pi_j^2}
$$

When $\pi_j = 1/k$ for all $j$, this reduces to the JC($k$) model
(a.k.a. Mk).

### Dirichlet-marginal discretisation

Rather than sampling full frequency vectors from
$\mathrm{Dirichlet}(\alpha, \ldots, \alpha)$, we exploit the fact that
**the marginal distribution of any single component** is:

$$
\pi_s \sim \mathrm{Beta}(\alpha, \; (k-1)\alpha)
$$

This is a standard property of the Dirichlet distribution. It allows us
to reduce a $k$-dimensional integral to a one-dimensional discretisation
that adapts automatically to the number of states $k$.

We discretise $\mathrm{Beta}(\alpha, (k-1)\alpha)$ into $B$
equal-probability bins (default $B = 4$). The representative value for
bin $b$ is the **midpoint of the conditional expectation** within that
bin:

$$
\beta_b = \frac{
  F^{-1}_{\mathrm{Beta}}\!\bigl(\tfrac{b - 0.5}{B}\bigr)
}{}
$$

where $F^{-1}_{\mathrm{Beta}}$ is the quantile function of
$\mathrm{Beta}(\alpha, (k-1)\alpha)$.

More precisely, $\beta_b$ is the quantile at probability $(2b - 1) / (2B)$,
the midpoint of the $b$-th equal-probability interval.

### Frequency vectors via rotation

Given a bin value $\beta_b$, we construct $k$ **rotated** frequency
vectors, one for each "focal" state $s \in \{0, \ldots, k-1\}$:

$$
\pi^{(b,s)}_j = \begin{cases}
  \beta_b & \text{if } j = s \\[4pt]
  \dfrac{1 - \beta_b}{k - 1} & \text{if } j \neq s
\end{cases}
$$

The rotation serves two purposes:

1. **State-labelling symmetry.** Morphological state labels (0, 1, 2, ...)
   are arbitrary. Averaging over all $k$ rotations ensures that no state
   is systematically favoured.
2. **Correct marginal mean.** The average frequency vector across
   rotations is always $(1/k, \ldots, 1/k)$, regardless of $\beta_b$.

### Likelihood integration

The per-character likelihood under heterogeneity is the average over
all bins and rotations:

$$
L_c \;\approx\; \frac{1}{B} \cdot \frac{1}{k}
  \sum_{b=1}^{B} \sum_{s=0}^{k-1} L_c\!\bigl(\boldsymbol{\pi}^{(b,s)}\bigr)
$$
where $L_c(\boldsymbol{\pi})$ denotes the standard Felsenstein pruning
likelihood for character $c$ evaluated under the F81 model with
equilibrium frequencies $\boldsymbol{\pi}$.

#### Binary characters ($k = 2$)

For $k = 2$, $\mathrm{Beta}(\alpha, \alpha)$ is symmetric about $1/2$.
By symmetry, the two rotations for each bin produce the same likelihood
(the focal state is swapped, but F81 only cares about the frequency
partition). MkPrime therefore uses only one rotation per bin for binary
characters, halving the computation.

### Composition with among-character rate variation

Q-matrix heterogeneity and among-character rate variation (ACRV) are
independent and compose multiplicatively. When both are enabled, each
character's likelihood is averaged over $n_{\mathrm{Cat}} \times B \times
n_{\mathrm{rot}}$ components, where $n_{\mathrm{rot}} = 1$ for $k = 2$
and $n_{\mathrm{rot}} = k$ for $k \geq 3$.

### Composition with `rate_loss` (neomorphic characters)

Neomorphic characters use an asymmetric binary model parameterised by
`rate_loss` ($\rho$), where the gain rate is $1/(1 + \rho)$ and the loss
rate is $\rho/(1 + \rho)$, normalised so the total rate is 1.

Under heterogeneity, the equilibrium frequency for bin $b$ modulates
the asymmetry. The effective gain and loss rates become:

$$
\text{gain}_b = \frac{1}{1 + \rho} \cdot 2\beta_b, \qquad
\text{loss}_b = \frac{\rho}{1 + \rho} \cdot 2(1 - \beta_b)
$$

This is algebraically equivalent to an F81 model with
$\pi_1 = \text{gain}_b / (\text{gain}_b + \text{loss}_b)$.
The `rate_loss` parameter retains its interpretation as the global
asymmetry between gains and losses; `beta_scale` modulates how much
that asymmetry can vary across characters.

## The `beta_scale` parameter

`beta_scale` ($= \alpha$) is the single scalar controlling all
heterogeneity. Its interpretation:

| $\alpha$ | Behaviour |
|----------|-----------|
| $\alpha \to \infty$ | All bins concentrate at $1/k$; recovers JC($k$). No heterogeneity. |
| $\alpha = 1$ | Uniform distribution on the simplex. Moderate heterogeneity. |
| $\alpha \to 0$ | Bins spread to the boundaries; extreme heterogeneity (one state dominates entirely). |

The MCMC places a $\mathrm{Gamma}(\text{shape}, \text{rate})$ prior on
`beta_scale` and proposes new values via a scale move (multiplicative
random walk).

## Computational cost

The number of likelihood evaluations per character per MCMC step is
multiplied by:

- **Binary characters ($k = 2$):** $B$ (no rotation needed).
- **Multistate characters ($k \geq 3$):** $B \times k$.

For a typical morphological dataset that is 80--90% binary with
$B = 4$, the effective cost increase is roughly $4$--$5\times$.

## Approximation quality

The "one-special-state" frequency vector $\boldsymbol{\pi}^{(b,s)}$
cannot represent arbitrary Dirichlet draws. For $k = 3$, a true draw
might be $(0.5, 0.3, 0.2)$, which is not expressible as any rotation
of $(\beta, (1-\beta)/2, (1-\beta)/2)$.

This is inherent to any finite discretisation of a multivariate
distribution via its marginal. The approximation captures:

- The **marginal** variation in individual state frequencies (exactly).
- The **mean** frequency vector (exactly: always $1/k$ by rotation
  symmetry).
- The dominant mode of variation: one state elevated, the rest
  depressed equally.

It does **not** capture simultaneous elevation of two states at the
expense of others. For the Mk model—where state labels are arbitrary
and carry no biological meaning—this is an acceptable trade-off.
The primary signal is "some states are more common than others," not
"states 0 and 1 are both common while state 2 is rare."

## Comparison with RevBayes multistate approach

| Aspect | MkPrime | RevBayes multistate |
|--------|---------|---------------------|
| Parameters | 1 scalar (`beta_scale`) | $n_{\mathrm{cats}} \times k$ frequency vectors + mixture weights per $k$ |
| Mixture components | $B \times k$ per $k$ (deterministic given $\alpha$) | $2 \times n_{\mathrm{cats}}$ per $k$ (stochastic) |
| Symmetry | Rotation averaging | Reversed copies |
| MCMC mixing | Fast (1D scale move) | Slower (simplex moves in high dimensions) |
| Flexibility | Moderate (marginal discretisation) | High (arbitrary frequency vectors) |

The MkPrime approach trades some flexibility for dramatically better
MCMC mixing and a simpler parameterisation. For model comparison
purposes—where the goal is "does the data support unequal
frequencies?"—the single-parameter design is sufficient.

## Usage

```{r het-example}
#| eval: false
model <- MkPrimeModel(
  qHeterogeneity = TRUE,
  nBetaCat       = 4L,
  betaScaleShape = 1,
  betaScaleRate  = 1
)
```

Compare with a standard (no-Het) analysis using stepping-stone marginal
likelihoods:

```{r het-comparison}
#| eval: false
# Standard model
resultStd <- RunMkPrime(data, tree, model = MkPrimeModel())

# Het model
resultHet <- RunMkPrime(data, tree, model = MkPrimeModel(qHeterogeneity = TRUE))

# Compare via stepping-stone
ssStd <- mkp_stepping_stone(data, tree, model = MkPrimeModel())
ssHet <- mkp_stepping_stone(data, tree, model = MkPrimeModel(qHeterogeneity = TRUE))
```

The model with the higher log marginal likelihood is preferred.

## References
