---
title: "Bayesian Mk' analysis of the hyolith dataset"
format: html
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Bayesian Mk' analysis of the hyolith dataset}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library(MkPrime)
```

## Overview

This vignette walks through a complete Bayesian phylogenetic analysis using
MkPrime on a real morphological dataset: 225 characters scored for 54 hyolith
and outgroup taxa, from @Sun2018, available in the **TreeSearch**
package as `Sun2018.nex`.

The Mk' model extends the standard Mk model by treating the true number of
character states as a latent variable. Rather than assuming that all states
seen in the data are the only states that exist, Mk' infers a per-character k'
from a shared hyperprior. This matters for morphological data where characters
are scored for the organisms currently in hand and unobserved states are
plausible.

## Packages

All steps below use packages already listed in MkPrime's dependencies:
**ape**, **TreeTools**, and **TreeSearch** (for the example dataset).

```{r packages}
#| message: false
library(MkPrime)
```

## Loading morphological data

Read the NEXUS file with `TreeTools::ReadAsPhyDat()`, which correctly
handles MorphoBank-style NEXUS files including polymorphic state codings
(`(0,1)`, `{1,2}`, etc.):

```{r load-data}
nexFile <- system.file("datasets/Sun2018.nex", package = "TreeSearch")
pd <- TreeTools::ReadAsPhyDat(nexFile)
cat(sprintf(
  "phyDat: %d taxa, %d characters, %d state levels\n",
  length(pd),
  sum(attr(pd, "weight")),
  length(attr(pd, "levels"))
))
```

## Classifying characters

MkPrime supports three character types in a single analysis:

| Type | Argument | When to use |
|------|----------|-------------|
| **Neomorphic** | `neomorphic =` | Binary absent/present; asymmetric gain/loss rates |
| **Transformational** | (default) | True k' unknown and inferred from the data |
| **Known state space** | `knownStates =` | k fixed and known *a priori* |

This NEXUS file encodes character type in the `STATELABELS` block: characters
whose first state label is `[Transformational character]` are ordered
transformational series; the remaining binary characters are absent/present
neomorphic characters. We parse this directly:

```{r parse-character-types}
nexLines <- readLines(nexFile)

# Locate the STATELABELS and MATRIX blocks
slStart    <- grep("^\\s+STATELABELS", nexLines) + 1L
matrixLine <- grep("^\\s+MATRIX",      nexLines)
slEnd      <- matrixLine[matrixLine > slStart][1L] - 1L
slBlock    <- nexLines[slStart:slEnd]

# Lines containing only a number introduce each character's label set
charHeaderLines <- grep("^\\s+\\d+\\s*$", slBlock)
charNums         <- as.integer(trimws(slBlock[charHeaderLines]))

# Collect all state labels per character; test for the marker phrase
charLabelRanges <- Map(
  function(start, end) slBlock[start:end],
  charHeaderLines + 1L,
  c(charHeaderLines[-1L] - 1L, length(slBlock))
)
isTrans <- vapply(charLabelRanges, function(lbls) {
  any(grepl("\\[Transformational character\\]", lbls, ignore.case = TRUE))
}, logical(1))

transformationalChars <- charNums[isTrans]
neomorphicChars       <- charNums[!isTrans]

cat(sprintf(
  "%d transformational, %d neomorphic characters\n",
  length(transformationalChars), length(neomorphicChars)
))
```

Characters not listed in `neomorphic` default to transformational in
`MkPrimeData()`, so we only need to pass the neomorphic indices.

## Preparing the data object

`MkPrimeData()` validates and partitions the characters:

```{r mkprime-data}
mkd <- MkPrimeData(pd, neomorphic = neomorphicChars)
mkd
```

## Model specification

`MkPrimeModel()` configures priors and the statistical model:

```{r model-spec}
model <- MkPrimeModel(
  coding  = "variable",  # @Lewis2001 ascertainment bias correction
  nCat    = 6L,          # discrete lognormal ACRV rate categories
  relabel = TRUE         # Mk' relabelling correction for transformational chars
)
model
```

`expSteps = NULL` (default) sets the tree-length prior scale automatically
from the Fitch parsimony score of the starting tree.

### Q-matrix heterogeneity (optional)

Standard Mk assumes all characters share equal state frequencies. To relax
this, enable Q-matrix heterogeneity, which integrates each character's
likelihood over a mixture of F81 rate matrices with varying equilibrium
frequencies controlled by a single parameter, `beta_scale`:

```{r model-het}
#| eval: false
modelHet <- MkPrimeModel(
  coding           = "variable",
  nCat             = 6L,
  qHeterogeneity   = TRUE,
  nBetaCat         = 4L        # number of discretisation bins
)
```

This is primarily useful for **model comparison**: run analyses with and
without heterogeneity, then compare marginal likelihoods via
`mkp_stepping_stone()` to see whether the data support unequal state
frequencies. Expect roughly 4--5$\times$ longer run times for
binary-dominated datasets. See `vignette("het-details")` for the
mathematical details.

## MCMC configuration

MkPrime uses a three-phase MCMC architecture:

1. **Warmup** — step sizes and move weights adapt; chain stabilisation is
   detected automatically via a Geweke z-score test on the log-posterior.
   No samples are saved.
2. **Tuning** — move weights are optimised to maximise the minimum effective
   sample size per second (min-ESS/s) using a perturbation bandit.
   Samples are discarded.
3. **Sampling** — move weights are frozen and posterior samples are collected
   until convergence criteria are met.

For a **production run** on this dataset we recommend:

```{r mcmc-production}
mcmcProduction <- MkPrimeMCMC(
  # nIter = Inf (default): run until convergence criteria are met
  # thin = "auto" (default): one sample per full move cycle
  nRuns          = 2L,       # 2 independent runs (required for R-hat)
  nChains        = 4L,       # parallel tempering chains
  heat           = 0.2,
  minEss         = 200L,     # stop when min ESS > 200
  maxRhat        = 1.01,     # stop when max R-hat < 1.01
  checkEvery     = 5000L,
  logFile        = "hyoliths.log"
)
```

## Running the analysis

Pass the `phyDat` object directly to `RunMkPrime()` together with the
neomorphic character indices. The function calls `MkPrimeData()` internally if
you pass a `phyDat` rather than a pre-built `MkPrimeData` object. A
neighbour-joining starting tree is built automatically from the data;
pass `tree =` explicitly if you prefer a different starting topology.

MCMC options can be passed inline or via a pre-built `MkPrimeMCMC` object.
For the production run we use the object defined above; for the demo run we
pass the settings directly.

Production run (not evaluated here — expected wall time: 1–4 hours):

```{r run-production}
#| eval: false
set.seed(8203)
posterior <- RunMkPrime(
  data       = pd,
  neomorphic = neomorphicChars,
  model      = model,
  mcmc       = mcmcProduction
)
saveRDS(posterior, "hyoliths_posterior.rds")
```

Demo run:

```{r run-demo}
#| cache: true
set.seed(8203)
posteriorDemo <- RunMkPrime(
  data       = pd,
  neomorphic = neomorphicChars,
  model      = model,
  nIter      = 5000L,
  thin       = 10L,
  maxWarmup  = 2000L,
  autoTune   = FALSE,
  nRuns      = 2L,
  nChains    = 1L
)
```

## Convergence diagnostics

`ConvergenceDiagnostics()` computes the effective sample size (ESS) and
rank-normalized split-R-hat (Vehtari et al. 2021) across runs.
R-hat supersedes the classical PSRF (Gelman--Rubin), which assumes
approximate normality:

```{r convergence}
diag <- ConvergenceDiagnostics(posterior)
print(diag)
```

A well-converged run should show ESS ≥ 200 and R-hat ≤ 1.01 for all scalar
parameters. The demo run is far too short to meet these criteria.

Trace plots give a visual check on stationarity and mixing:

```{r trace-plot}
#| fig-width: 7
#| fig-height: 5
plot(posteriorDemo)
```

## Summarizing the posterior

`summary()` reports posterior mean, median, 95% credible interval, ESS, and
R-hat for each scalar parameter:

```{r summary}
summary(posteriorDemo)
```

### Key parameters

| Parameter | Interpretation |
|-----------|---------------|
| `log_posterior` | Log unnormalized posterior density |
| `tree_length` | Total branch length |
| `rate_loss` | Asymmetry for neomorphic characters (rate loss / rate gain) |
| `rate_log_sd` | Dispersion of the lognormal ACRV distribution |
| `p` | Geometric hyperprior parameter; controls prior expected k' |
| `kPrime_[i]` | Posterior k' for transformational character *i* |

### k' posterior estimates

Characters with posterior k' > kObs (observed state count) carry probability
that additional states exist but were not sampled. To examine these from a
production run:

```{r kprime-summary}
kprimeCols <- grep("^kPrime_", colnames(posteriorDemo$samples), value = TRUE)
if (length(kprimeCols) > 0L) {
  kpMeans <- colMeans(posteriorDemo$samples[, kprimeCols, drop = FALSE])
  head(sort(kpMeans, decreasing = TRUE), 10)
}
```

### Posterior tree distribution

Sampled trees are stored in the returned object.

```{r tree-summary}
#| eval: false
# Majority-rule consensus from a production run
# posterior$trees already combines all runs; apply extra burnin if desired:
# posterior <- SetBurnin(posterior, 500)
allTrees <- posterior$trees

# Identify and exclude rogue taxa that destabilize the consensus
rogues <- Rogue::QuickRogue(allTrees)
rogueNames <- rogues$taxon[nzchar(rogues$taxon, keepNA = FALSE)]
if (length(rogueNames) > 0L) {
  cat("Rogue taxa identified:", paste(rogueNames, collapse = ", "), "\n")
  allTrees <- lapply(allTrees, ape::drop.tip, tip = rogueNames)
  class(allTrees) <- "multiPhylo"
}

consensus <- TreeTools::Consensus(allTrees, p = 0.5)
par(mar = rep(0, 4))
outgroup <- c("Tonicella", "Serpula", "Loxosomella")
par(mar = rep(0, 4))
plot(TreeTools::SortTree(TreeTools::RootTree(consensus, outgroup)),
     cex = 0.6, main = "Majority-rule consensus")
```


```{r tree-summary-demo}
#| eval: true
# Majority-rule consensus from a demo run
allTrees <- posteriorDemo$trees[-(1:100)]


# Identify rogue taxa
rogues <- Rogue::QuickRogue(sample(allTrees, 10))
rogueNames <- rogues$taxon[nzchar(rogues$taxon, keepNA = FALSE)]
if (length(rogueNames) > 0L) {
  cat("Rogue taxa:", paste(rogueNames, collapse = ", "), "\n")
  allTrees <- lapply(allTrees, ape::drop.tip, tip = rogueNames)
  class(allTrees) <- "multiPhylo"
}

consensus <- TreeTools::Consensus(allTrees, p = 0.5)
consTitle <- if (length(rogueNames) > 0L) "Majority-rule consensus (rogues excluded)" else "Majority-rule consensus"
par(mar = rep(0, 4))
plot(consensus, cex = 0.6, main = consTitle)
```


## Saving and resuming

```{r save-load}
#| eval: false
saveRDS(posterior, "hyoliths_posterior.rds")
posterior <- readRDS("hyoliths_posterior.rds")
```

To enable checkpoint files (which allow resuming after interruption):

```{r checkpoint}
#| eval: false
mcmcCheckpointed <- MkPrimeMCMC(
  nIter          = 200000L,
  nRuns          = 2L,
  nChains        = 4L,
  checkpointFile = "hyoliths_checkpoint.rds"
)
```

