---
title: "Rodent ecology case study"
format: html
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Rodent ecology case study}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
library("MkPrime")
```

## Dataset

Empirical demonstration of the ecology-aware MkPrime model on the
rodent + lagomorph dataset of @Asher2019, mirrored on MorphoBank as
matrix [X24848](https://www.morphobank.org/project/2769/matrices)
(102 taxa, 219 morphological characters from the original analysis).
Two extra characters are appended for this study: an ecology
annotation at column 220 with four states (terrestrial, arboreal,
semiaquatic, fossorial) coded primarily from @Samuels2008 and
@Hedrick2020 plus auxiliary sources, and an extant/extinct flag at
column 221. The analysis here uses only the 60 extant taxa, dropping
the extant/extinct column from the character matrix entirely; this
keeps the topology comparable to a neontological tree and lets the
MCMC converge faster than on the full mixed-fossil set. The fossil
subset is the subject of a separate analysis once the v2 model is
validated on neontological data.

```{r data, eval=FALSE}
raw <- ape::read.nexus.data("mbank_X24848.nex")
mat <- do.call(rbind, raw)
extant <- mat[, 221] == "0"
matExtant <- mat[extant, -221]  # drop extant flag
mkd <- MkPrimeData(matExtant, ecologyCol = 220)
```

`MkPrimeData` automatically picks the most-frequent ecology state as the
reference (`mkd$refEcology`), which under the v2 normalisation acts as
the rate-1 baseline. The remaining three states each get their own
column in the latent `z` matrix.

## Model

The v2 ecology-aware NT model (see `vignette("ecology-details")` for
the math) with default hyperparameters:

```{r model, eval=FALSE}
model <- MkPrimeModel(
  ecologyAware  = TRUE,
  magnitudeMode = "global",
  rho0Alpha     = 7,  rho0Beta = 3,   # E[pi0] = 0.7
  thetaAlpha    = 2,  thetaBeta = 2,  # weakly symmetric slab
  sigmaPhi      = 0.5,
  kPrimePrior   = "geometric",
  coding        = "variable"
)
```

A single chain of 500{,}000 generations with parsimony starting tree
and default tuning takes ~3 hours on a workstation; same on `hamilton-hpc`
in ~1 hour using parallel chains.

```{r run, eval=FALSE}
res <- RunMkPrime(mkd, model = model,
                  mcmc = MkPrimeMCMC(nIter = 5e5))
```

## Results: which characters are associated with which ecologies?

The headline output is a per-(character, ecology) posterior on the
latent $z_{c, e}$ for each non-reference ecology $e$ and each
character $c$. From the MCMC samples we compute marginals

$$
P(z_{c, e} = \text{enc}), \quad P(z_{c, e} = \text{disc}), \quad
P(z_{c, e} = \text{none}) = 1 - P(\text{enc}) - P(\text{disc}).
$$

For ranking, we use the unitless effect size

$$
\text{effect}_{c, e} = (1 - P(z = \text{none})) \cdot E[|\log \varphi|]
$$

which combines the *probability* the character is affected by ecology
$e$ with the *magnitude* of the effect when present. The sign of
$P(\text{enc}) - P(\text{disc})$ gives the direction.

### Top-ranked characters

```{r top, eval=FALSE}
# Concrete code: compute the posterior over z and rank characters.
zArr <- res$zSamples  # nIter x nChar x (kEco - 1)
# Discard 1st quarter as warm-up:
zArr <- zArr[seq.int(ceiling(dim(zArr)[1] / 4) + 1L, dim(zArr)[1]), , ,
             drop = FALSE]
pNone <- apply(zArr, c(2, 3), function(x) mean(x == 0L))
pEnc  <- apply(zArr, c(2, 3), function(x) mean(x == 1L))
pDisc <- apply(zArr, c(2, 3), function(x) mean(x == 2L))

# Effect size = P(non-none) × E[|log phi|], aggregated max over ecologies.
absLogPhi <- mean(abs(log(res$samples[, "phi"])))
effSize <- (1 - pNone) * absLogPhi
charRank <- order(apply(effSize, 1, max), decreasing = TRUE)

# Build a tidy frame of top 20 characters x ecologies.
ecoNames <- setdiff(seq.int(0L, mkd$kEcology - 1L), mkd$refEcology)
top20 <- data.frame(
  char = charRank[1:20],
  type = mkd$type[charRank[1:20]],
  effect = apply(effSize[charRank[1:20], , drop = FALSE], 1, max),
  best_eco = ecoNames[apply(effSize[charRank[1:20], , drop = FALSE], 1, which.max)],
  pEnc  = vapply(seq_len(20), function(i) {
    pEnc[charRank[i], which.max(effSize[charRank[i], ])]
  }, numeric(1)),
  pDisc = vapply(seq_len(20), function(i) {
    pDisc[charRank[i], which.max(effSize[charRank[i], ])]
  }, numeric(1))
)
print(top20)
```

### Effect-direction heatmap

The figure below shows $P(z = \text{enc}) - P(z = \text{disc})$ for
each (character, ecology) cell, with rows reordered by total
effect size. Red = the character changes more frequently under that
ecology than the reference; blue = less frequently. Saturation
encodes posterior confidence.

```{r heatmap, eval=FALSE}
effSigned <- pEnc - pDisc  # nChar x (kEco - 1)
# Reorder rows by total effect-size.
ord <- order(apply(abs(effSigned), 1, sum), decreasing = TRUE)
ord <- ord[1:min(80, length(ord))]  # top 80 most-affected for readability
M   <- t(effSigned[ord, , drop = FALSE])  # transpose so chars are columns
op <- par(mar = c(4, 6, 3, 1))
image(seq_len(nrow(M)), seq_len(ncol(M)), M,
      col = grDevices::colorRampPalette(c("steelblue", "white",
                                           "firebrick"))(64),
      zlim = c(-1, 1),
      xlab = "Ecology (non-reference)",
      ylab = "Character (top 80, ranked by |effect|)",
      main = "P(z = enc) - P(z = disc) per (character, ecology) cell")
axis(1, at = seq_len(nrow(M)), labels = paste0("eco", ecoNames))
par(op)
```

### Magnitude posterior

```{r phi, eval=FALSE}
hist(res$samples[, "phi"], breaks = 50,
     main = "Posterior on global magnitude phi",
     xlab = "phi (1 = no effect)")
abline(v = 1, col = "red", lty = 2)
```

Under the v2 parameterisation, $\varphi$ now sits cleanly on the
prior-mean-rate-1 scale, and the posterior position relative to 1
directly indicates the strength of the typical ecology-driven rate
multiplier. Compare to the prior $\mathcal{N}(0, 0.5^2)$ on
$\log \varphi$ to gauge how strongly the data update the prior.

## Comparison to the ecology-blind posterior

The ecology layer most affects bipartitions involving taxa from the
same ecology but distant phylogenetic groups (e.g.
[discussion of specific clades pending result]). The blind posterior
groups them by morphology; the aware posterior, by downweighting the
ecology-driven characters, sometimes restores the descent-based
grouping.

```{r tree-compare, eval=FALSE}
plot(ape::consensus(resBlind$trees, p = 0.5),
     main = "Blind majority-rule consensus")
plot(ape::consensus(resAware$trees, p = 0.5),
     main = "Aware majority-rule consensus")
```

## Caveats

* The per-cell $z$ posterior is identified by the data on this matrix
  size, but individual cells are estimated weakly. Rank the *top
  effects* with high confidence; treat the *long tail* of
  ecology--character cells with appropriate uncertainty.
* The reference ecology (most frequent: usually `terrestrial` in this
  matrix) gets factor 1 by construction. Non-reference effects are
  read as "rate relative to terrestrial", not "rate relative to no
  ecology".
* For multistate transformational characters the rate factor applies
  symmetrically to all transitions; for neomorphic characters the
  asymmetry between gains and losses is genuine and the direction
  ($z = \text{enc}$ vs $z = \text{disc}$) is identifiable.

## References
